Where Do Events Come From?

### 2018-08-05 paper.tex
WIP

Reactive applications interact continuously and in real time with the external
world through hardware peripherals such as sensors and actuators (e.g.,
buttons, displays, timers, etc.).
%
These interactions are typically represented in software as input events
flowing from the peripherals to the application and as output events flowing
from the application to the peripherals.
%
As illustrated in Figure~\ref{fig.env}, peripherals can be encapsulated as a
single component, \emph{the environment}, which connects with the application
in an event loop:
the application sits idle until the environment awakes it on the occurrence of
an input;
the application reacts to the input and generates back one or more outputs
which affect the environment;
the application becomes idle and the loop restarts.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{loop}
\caption{ Event loop in reactive systems.
          The environment controls the application through input \& output
          events.
\label{fig.env}
}
\end{figure}

@@

Our work is largely inspired by TinyOS~\cite{wsn.tos}, a low-power OS for
wireless devices, which also provides asynchronous events with a compile-time
data race detector~\cite{wsn.nesc}, and automatic energy management triggered
from idle CPU states~\cite{wsn.icem}.
%
We adapted these ideas to the structured reactive programming model of \CEU,
and refined them with stronger guarantees due to lexical scope and automatic
finalization.

@@

Structured reactive programming lets developers write code in direct style,
recovering from the inversion of control imposed by event-driven
execution~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.

@@

The \code{par/or}, which stands for \emph{parallel-or}, is a lexical
composition that terminates as soon as one of the trails terminates, but which
also automatically aborts and finalizes the other trail(s).
%
In the example, when the button is pressed (ln~5), not only the toggling loop
will be aborted (ln~10--15), but the \code{finalize} clause will turn the LED
off unconditionally (ln~8), since its enclosing block goes out of scope.
%
The \code{par/or} is regarded as an \emph{orthogonal preemption
primitive}~\cite{esterel.preemption} because the trails need not to be tweaked
in order to affect each other.

The synchronous execution model of \CEU dictates that reactions to input events
are atomic and that incoming events are never lost.
%
The event loop in Figure~\ref{fig.env} suggests this behavior since the
environment can only generate a new input after the application yields control
and closes the loop.
%
In Listing~\ref{lst.ceu}, even if two packets arrive simultaneously in the
environment, the synchronous model ensures atomicity and responsiveness in the
application:
    the first \code{await} awakes and turns the LED off atomically
    (ln~12--13);
    and the second \code{await} will awake in sequence (ln~14).
%
%Note that if multiple trails in parallel await the same input, they will all
%awake from it during the same reaction.
%In this case, \CEU dictates that each trail executes atomically, one after the
%other, in lexical order.
%This rule prevents race conditions for trails in parallel.

Figure~\ref{fig.ticks} illustrates the synchronous execution model of \CEU.
%
The continuous timeline (representing the environment) shows
``physical timestamps'' for the event occurrences (e.g., event~\code{C}
occurs exactly at 17ms521us).
%
The discrete timeline in the application shows how the same occurring events
fit in the logical notion of time of \CEU.
%
The boot reaction \code{boot-0} happens before any input and starts the
program.
%
Event~\code{A} ``physically'' occurs during \code{boot-0} but, because time
is discrete, its corresponding reaction only executes at logical
instant~\code{A-1}.
%
Similarly, event~\code{B} occurs during~\code{A-1} and its reaction is
postponed to execute at~\code{B-2}.
%
Event~\code{C} also occurs during~\code{A-1} but its reaction must also wait
for~\code{B-2} to execute at~\code{C-3}.
%
Finally, event~\code{D} occurs during an idle period and can start immediately
at \code{D-4}.
%
%Finally, two instances of event~\code{E} occur during~\code{D-4}; they are
%handled in the subsequent reactions~\code{E-5} and~\code{E-6}.

\begingroup
\begin{figure}[t]
\centering
\begin{tikzpicture}[x=1em,y=1em,
  font=\footnotesize\sffamily,>=stealth,
  time/.style={font=\scriptsize\sffamily},
  cross/.style={inner sep=0pt,outer sep=0pt},
  evt/.style={fill=white,inner sep=1.85pt,outer sep=0pt},
  rng/.style={fill=white,inner sep=1pt,outer sep=0pt,
    font=\ttfamily\footnotesize}]
  \draw[->]
  (0,0)
    node[align=right,anchor=east,xshift=-.5em]
        {Continuous\\Timeline\\(Env.)}
    node[anchor=north west,xshift=-.5em,time]
        {0ms000us}
    node[anchor=north west,xshift=8.75em,time]
        {17ms521us}
  -- (18.5,0)
    node[anchor=north east,xshift=.1em,time]
        {30ms000us};
  \node[cross] at (0,0){x};
  %%
  \coordinate(Ext) at ($(0,0)!.5!(0,-5)$);
  \coordinate(A) at (Ext-|4,0);
  \coordinate(B) at (Ext-|5.5,0);
  \coordinate(C) at (Ext-|9,0);
  \coordinate(D) at (Ext-|15,0);
  %%
  \draw[semithick,dotted] (0,1-|A) -- (0,-6-|A);
  \draw[semithick,dotted] (0,1-|B) -- (0,-6-|B);
  \draw[semithick,dotted] (0,1-|C) -- (0,-6-|C);
  \draw[semithick,dotted] (0,1-|D) -- (0,-5-|D);
  \node[evt](Aevt) at (A) {\texttt{\textbf{A}}};
  \node[evt](Bevt) at (B) {\texttt{\textbf{B}}};
  \node[evt](Cevt) at (C) {\texttt{\textbf{C}}};
  \node[evt](Devt) at (D) {\texttt{\textbf{D}}};
  \node[anchor=east,xshift=-.5em] at (A){\emph{External Input}};
  %%
  \draw[-]
  (0,-5)
    node[align=right,anchor=east,xshift=-.5em]
        {Discrete\\Timeline\\(App.)}
  -- (18.5,-5);
  \node[cross](P0) at (0,-5){x};
  \node[cross](P1) at (4.75,-5){x};
  \node[cross](P2) at (10,-5){x};
  \node[cross](P3) at (12.5,-5){x};
  \node[cross](P4) at (14.5,-5){x};
  \node[cross](P5) at (15.5,-5){x};
  \node[cross](P6) at (18.5,-5){x};
  \node[rng] at ($(P0)!.5!(P1)$){boot-0};
  \node[rng] at ($(P1)!.5!(P2)$){A-1};
  \node[rng] at ($(P2)!.5!(P3)$){B-2};
  \node[rng] at ($(P3)!.5!(P4)$){C-3};
  \node[rng] at ($(P5)!.5!(P6)$){D-4};
  \draw[decorate,decoration={brace,raise=-1pt,mirror,amplitude=2pt}]
  ($(P4)+(0,-.5)$) -- node[below,time] {Idle period} ($(P5)+(0,-.5)$);
  %%
  \begin{scope}[shorten >=1.5pt]
  \draw[->] (Aevt) -- (P1);
  \draw[->] (Bevt) -- (P2);
  \draw[->] (Cevt) -- (P3);
  \draw[->] (Devt) -- (P5);
  \end{scope}
\end{tikzpicture}
\vskip-.6em
%\includegraphics[width=\columnwidth]{tick_min}
\caption{The discrete notion of time in \CEU.}
\label{fig.ticks}
\end{figure}
\endgroup

In order to guarantee responsiveness, the synchronous model relies on the
\emph{synchronous hypothesis}, which states that reactions must be
significantly faster than the rate of inputs.
%
For this reason, \CEU (like most synchronous languages) refuses unbounded loops
at compile time.
This restriction guarantees that all reactions to the environment are computed
in bounded time~\cite{ceu.lctes18}, ensuring that applications are always
responsive to upcoming events.

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{sync-async}
\caption{ An application in \CEU has a concurrent asynchronous side and a
          predictable synchronous side that receives and reacts to inputs
          atomically.
\label{fig.async}
}
\end{figure}

We propose to extend \CEU with asynchronous ISRs.
However, asynchronous execution confronts the synchronous mindset of \CEU since
the assumption that reactions are atomic no longer holds.
Not only this may lead to race conditions at a fine grain, but may also affect
the ordering of events at a coarse grain: the effect of an earlier event may be
perceived after the effect of a later event.
%
Still, our goal is to preserve the well-behaved interaction between the \CEU
application and the environment even in the presence of asynchronous ISRs.
Figure~\ref{fig.async} adapts the event loop of Figure~\ref{fig.env}, with the
original application represented as the synchronous side and the environment as
the asynchronous side, all inside the same \CEU application.
%
Our idea is to push all subtleties of asynchronous execution into device
drivers, which emit input events to regular synchronous code in \CEU, exactly
as before.

@@

The example illustrates the clear separation between the application and the
driver through an event-driven interface that reflects the desired architecture
of Figures~\ref{fig.env}~and~\ref{fig.async}.
%
Now, the whole application is written in \CEU:
    the synchronous side remains well behaved with no low-level calls, and
    the asynchronous side deals with the complexity of device drivers.
%
Nevertheless, drivers are typically write-once components developed by embedded
specialists, which are reused in most applications.
%
Note that each supported architecture requires a mapping between the
\code{async/isr} and the actual interrupt vector table (which we provide
for the \emph{AVR/ATmega} and \emph{ARM/Cortex} microcontrollers).

@@

\begin{figure}[t!]
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[
  numbers=right,
  xrightmargin=3em,
  basicstyle=\ttfamily\small,
  label={lst.mem.a},
  caption={Shared \code{x}},
  escapechar=\%,
]
input void A;
input void B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\end{minipage}%
%
\begin{minipage}[t]{0.50\linewidth}
%\begin{lstlisting}
\begin{lstlisting}[
  xleftmargin=3em,
  basicstyle=\ttfamily\small,
  label={lst.mem.b},
  caption={Shared \code{y}},
  escapechar=\%,
]
input void A;
// (empty line)
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end

@@

Our runtime implements the event loop proposed in Figure~\ref{fig.async}, which
alternates between synchronous, well-behaved execution, and asynchronous,
unpredictable ISRs.
%
Since the language now has full control over the event loop and only executes
from interrupts, it is possible for the runtime to enter in sleep mode
automatically to save energy.

@@

Note that applications need not to be explicit about energy management to take
advantage of sleep modes.
All happens automatically because of the synchronous-reactive execution model
and energy-aware runtime of \CEU.
Figure~\ref{fig.energy} shows our initial tests with five existing C/Arduino
applications rewritten in \CEU.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{energy}
\caption{ Energy consumption in \emph{mA} for five existing C/Arduino
          applications rewritten in \CEU.
\label{fig.energy}
}
\end{figure}

\section{Related Work}

@@

\begin{comment}
- CRP

@@

- electre
\end{comment}

>>>>

Reactive applications interact continuously and in real time with the external
world through hardware peripherals such as sensors and actuators (e.g.,
buttons, displays, timers, etc.).
%
These interactions are typically represented in software as input events
flowing from the peripherals to the application and as output events flowing
from the application to the peripherals.
%
%As illustrated in Figure~\ref{fig.env},
Peripherals can be encapsulated as a
single component, \emph{the environment}, which connects with the application
in an event loop:
the application sits idle until the environment awakes it on the occurrence of
an input;
the application reacts to the input and generates back one or more outputs
which affect the environment;
the application becomes idle and the loop restarts.

\begin{comment}
\begin{figure}
\centering
\includegraphics[width=\linewidth]{loop}
\caption{ Event loop in reactive systems.
          The environment controls the application through input \& output
          events.
\label{fig.env}
}
\end{figure}
\end{comment}

@@

\begin{comment}
Our work is largely inspired by TinyOS~\cite{wsn.tos}, a low-power OS for
wireless devices, which also provides asynchronous events with a compile-time
data race detector~\cite{wsn.nesc}, and automatic energy management triggered
from idle CPU states~\cite{wsn.icem}.
%
We adapted these ideas to the structured reactive programming model of \CEU,
and refined them with stronger guarantees due to lexical scope and automatic
finalization.
\end{comment}

@@

\begin{comment}
Structured reactive programming lets developers write code in direct style,
recovering from the inversion of control imposed by event-driven
execution~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
\end{comment}

@@

The \code{par/or}, which stands for \emph{parallel-or}, is a lexical
composition that terminates as soon as one of the trails terminates, but which
also automatically aborts and finalizes the other trail(s).
%
In the example, when the button is pressed (ln~5), not only the toggling loop
will be aborted (ln~10--15), but the \code{finalize} clause will turn the LED
off unconditionally (ln~8), since its enclosing block goes out of scope.
%
%The \code{par/or} is regarded as an \emph{orthogonal preemption
%primitive}~\cite{esterel.preemption} because the trails need not to be tweaked
%in order to affect each other.

The synchronous execution model of \CEU dictates that reactions to input events
are atomic and that incoming events are never lost.
%
%The event loop in Figure~\ref{fig.env} suggests this behavior since
The
environment can only generate a new input after the application yields control
and closes the event loop.
%
In Listing~\ref{lst.ceu}, even if two packets arrive simultaneously in the
environment, the synchronous model ensures atomicity and responsiveness in the
application:
    the first \code{await} awakes and turns the LED off atomically
    (ln~12--13);
    and the second \code{await} will awake in sequence (ln~14).
%
%Note that if multiple trails in parallel await the same input, they will all
%awake from it during the same reaction.
%In this case, \CEU dictates that each trail executes atomically, one after the
%other, in lexical order.
%This rule prevents race conditions for trails in parallel.

\begin{comment}
Figure~\ref{fig.ticks} illustrates the synchronous execution model of \CEU.
%
The continuous timeline (representing the environment) shows
``physical timestamps'' for the event occurrences (e.g., event~\code{C}
occurs exactly at 17ms521us).
%
The discrete timeline in the application shows how the same occurring events
fit in the logical notion of time of \CEU.
%
The boot reaction \code{boot-0} happens before any input and starts the
program.
%
Event~\code{A} ``physically'' occurs during \code{boot-0} but, because time
is discrete, its corresponding reaction only executes at logical
instant~\code{A-1}.
%
Similarly, event~\code{B} occurs during~\code{A-1} and its reaction is
postponed to execute at~\code{B-2}.
%
Event~\code{C} also occurs during~\code{A-1} but its reaction must also wait
for~\code{B-2} to execute at~\code{C-3}.
%
Finally, event~\code{D} occurs during an idle period and can start immediately
at \code{D-4}.
%
%Finally, two instances of event~\code{E} occur during~\code{D-4}; they are
%handled in the subsequent reactions~\code{E-5} and~\code{E-6}.

\begingroup
\begin{figure}[t]
\centering
\begin{tikzpicture}[x=1em,y=1em,
  font=\footnotesize\sffamily,>=stealth,
  time/.style={font=\scriptsize\sffamily},
  cross/.style={inner sep=0pt,outer sep=0pt},
  evt/.style={fill=white,inner sep=1.85pt,outer sep=0pt},
  rng/.style={fill=white,inner sep=1pt,outer sep=0pt,
    font=\ttfamily\footnotesize}]
  \draw[->]
  (0,0)
    node[align=right,anchor=east,xshift=-.5em]
        {Continuous\\Timeline\\(Env.)}
    node[anchor=north west,xshift=-.5em,time]
        {0ms000us}
    node[anchor=north west,xshift=8.75em,time]
        {17ms521us}
  -- (18.5,0)
    node[anchor=north east,xshift=.1em,time]
        {30ms000us};
  \node[cross] at (0,0){x};
  %%
  \coordinate(Ext) at ($(0,0)!.5!(0,-5)$);
  \coordinate(A) at (Ext-|4,0);
  \coordinate(B) at (Ext-|5.5,0);
  \coordinate(C) at (Ext-|9,0);
  \coordinate(D) at (Ext-|15,0);
  %%
  \draw[semithick,dotted] (0,1-|A) -- (0,-6-|A);
  \draw[semithick,dotted] (0,1-|B) -- (0,-6-|B);
  \draw[semithick,dotted] (0,1-|C) -- (0,-6-|C);
  \draw[semithick,dotted] (0,1-|D) -- (0,-5-|D);
  \node[evt](Aevt) at (A) {\texttt{\textbf{A}}};
  \node[evt](Bevt) at (B) {\texttt{\textbf{B}}};
  \node[evt](Cevt) at (C) {\texttt{\textbf{C}}};
  \node[evt](Devt) at (D) {\texttt{\textbf{D}}};
  \node[anchor=east,xshift=-.5em] at (A){\emph{External Input}};
  %%
  \draw[-]
  (0,-5)
    node[align=right,anchor=east,xshift=-.5em]
        {Discrete\\Timeline\\(App.)}
  -- (18.5,-5);
  \node[cross](P0) at (0,-5){x};
  \node[cross](P1) at (4.75,-5){x};
  \node[cross](P2) at (10,-5){x};
  \node[cross](P3) at (12.5,-5){x};
  \node[cross](P4) at (14.5,-5){x};
  \node[cross](P5) at (15.5,-5){x};
  \node[cross](P6) at (18.5,-5){x};
  \node[rng] at ($(P0)!.5!(P1)$){boot-0};
  \node[rng] at ($(P1)!.5!(P2)$){A-1};
  \node[rng] at ($(P2)!.5!(P3)$){B-2};
  \node[rng] at ($(P3)!.5!(P4)$){C-3};
  \node[rng] at ($(P5)!.5!(P6)$){D-4};
  \draw[decorate,decoration={brace,raise=-1pt,mirror,amplitude=2pt}]
  ($(P4)+(0,-.5)$) -- node[below,time] {Idle period} ($(P5)+(0,-.5)$);
  %%
  \begin{scope}[shorten >=1.5pt]
  \draw[->] (Aevt) -- (P1);
  \draw[->] (Bevt) -- (P2);
  \draw[->] (Cevt) -- (P3);
  \draw[->] (Devt) -- (P5);
  \end{scope}
\end{tikzpicture}
\vskip-.6em
%\includegraphics[width=\columnwidth]{tick_min}
\caption{The discrete notion of time in \CEU.}
\label{fig.ticks}
\end{figure}
\endgroup
\end{comment}

\begin{comment}
In order to guarantee responsiveness, the synchronous model relies on the
\emph{synchronous hypothesis}, which states that reactions must be
significantly faster than the rate of inputs.
%
For this reason, \CEU (like most synchronous languages) refuses unbounded loops
at compile time.
This restriction guarantees that all reactions to the environment are computed
in bounded time~\cite{ceu.lctes18}, ensuring that applications are always
responsive to upcoming events.
\end{comment}

@@

\begin{comment}
\begin{figure}
\centering
\includegraphics[width=\linewidth]{sync-async}
\caption{ An application in \CEU has a concurrent asynchronous side and a
          predictable synchronous side that receives and reacts to inputs
          atomically.
\label{fig.async}
}
\end{figure}
\end{comment}

We propose to extend \CEU with asynchronous ISRs.
However, asynchronous execution confronts the synchronous mindset of \CEU since
the assumption that reactions are atomic no longer holds.
Not only this may lead to race conditions at a fine grain, but may also affect
the ordering of events at a coarse grain: the effect of an earlier event may be
perceived after the effect of a later event.
%
Still, our goal is to preserve the well-behaved interaction between the \CEU
application and the environment even in the presence of asynchronous ISRs.
%Figure~\ref{fig.async} adapts the event loop of Figure~\ref{fig.env}, with the
%original application represented as the synchronous side and the environment as
%the asynchronous side, all inside the same \CEU application.
%
Our idea is to push all subtleties of asynchronous execution into device
drivers, which emit input events to regular synchronous code in \CEU, exactly
as before.

@@

The example illustrates the clear separation between the application and the
driver through an event-driven interface. % that reflects the desired architecture
%of Figures~\ref{fig.env}~and~\ref{fig.async}.
%
Now, the whole application is written in \CEU:
    the synchronous side remains well behaved with no low-level calls, and
    the asynchronous side deals with the complexity of device drivers.
%
Nevertheless, drivers are typically write-once components developed by embedded
specialists, which are reused in most applications.
%
Note that each supported architecture requires a mapping between the
\code{async/isr} and the actual interrupt vector table (which we provide
for the \emph{AVR/ATmega} and \emph{ARM/Cortex} microcontrollers).

@@

\begin{figure}[t]
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[
  numbers=right,
  xrightmargin=3em,
  basicstyle=\ttfamily\small,
  label={lst.mem.a},
  caption={Shared \code{x}},
  escapechar=\%,
]
input void A;
input void B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\end{minipage}%
%
\begin{minipage}[t]{0.50\linewidth}
%\begin{lstlisting}
\begin{lstlisting}[
  xleftmargin=3em,
  basicstyle=\ttfamily\small,
  label={lst.mem.b},
  caption={Shared \code{y}},
  escapechar=\%,
]
input void A;
// (empty line)
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end

@@

Our runtime
%implements the event loop proposed in Figure~\ref{fig.async}, which
alternates between synchronous, well-behaved execution, and asynchronous,
unpredictable ISRs.
%
Since the language now has full control over the event loop and only executes
from interrupts, it is possible for the runtime to enter in sleep mode
automatically to save energy.

@@

Note that applications need not to be explicit about energy management to take
advantage of sleep modes.
All happens automatically because of the synchronous-reactive execution model
and energy-aware runtime of \CEU.

\begin{comment}
Figure~\ref{fig.energy} shows our initial tests with five existing C/Arduino
applications rewritten in \CEU.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{energy}
\caption{ Energy consumption in \emph{mA} for five existing C/Arduino
          applications rewritten in \CEU.
\label{fig.energy}
}
\end{figure}
\end{comment}

\begin{comment}
\section{Related Work}

@@

%\begin{comment}
- CRP

@@

- electre
%\end{comment}
\end{comment}

\section{Conclusion}

Drivers

Our initial tests show optimal energy savings for applications in idle states.
